---
title: "Introduction to Stan and R Workflows"
author: "trinh dong"
date: "2025-11-23"
format:
    revealjs: 
        code-fold: true
        scroll-view: true
        scrollable: true

theme: simple
execute: 
    echo: true
---

## MCMC and HMC {.smaller}

- **MCMC (Markov chain Monte Carlo):** draws from posterior using a Markov chain.
- **HMC (Hamiltonian Monte Carlo):** efficient gradient-based MCMC used by Stan (fewer random-walk moves, better mixing for many models).

## What is Stan? {.smaller}

- Stan: probabilistic programming language implementing HMC/NUTS and other inference methods.
- Model specified via blocks: data, transformed data, parameters, transformed parameters, model, generated quantities.

## Stan program skeleton {.smaller}

```stan
// Demo stan code - Linear model
// trinhdhk
data {
    // metadata
    int<lower=0> N; // sample size
    int<lower=0> M; // feature dimension

    // design matrix
    vector[N] Y;    // outcome
    matrix[N, M] X; // model matrix, including intercept
}

parameters {
    vector[M] beta;      // coef
    real<lower=0> sigma; // dispersion term
}

transformed parameters {
   real lprior = 0.0; // if you want to investigate prior impact
   lprior += student_t_lpdf(beta | 3, 0, 2.5);
   lprior += student_t_lpdf(sigma | 3, 0, 2.5) - student_t_lccdf(0 | 3, 0, 2.5); //same as sigma ~ student_t(3,0,2.5)T[0, ];
}

model {
    vector[N] mu = X * beta; // quotation mark for transpose(X) and * is matmul.
    target += lprior;       
    target += normal_lpdf(Y | mu, sigma); 
}

generated quantities {
   vector[N] log_lik;
   vector[N] mu = X * beta; 
   for (n in 1:N){
        log_lik[n] = normal_lpdf(Y[n] | mu[n], sigma);
   }
} 
```

## Simulate data: linear regression  {.smaller}

```{r}
set.seed(123)
N <- 100
x <- rnorm(N, 0, 1)
alpha_true <- 1.0
beta_true <- 2.5
sigma_true <- 0.8
y <- alpha_true + beta_true * x + rnorm(N, 0, sigma_true)
df <- data.frame(x = x, y = y)
```

run this R code to create a toy dataset for the regression example. The variables `x` and `y` will be used as inputs to the Stan model.

## Prepare Stan data & compile model  {.smaller}

```{r}
library(cmdstanr)
# adjust path if your model is at `stan/LinearRegression.stan`
mod <- cmdstan_model("stan/LinearRegression.stan")
data_list <- list(N = N, M = 2, X = model.matrix(~x, data=df), Y = y)
```

## Fit with NUTS  {.smaller}

```{r}
fit <- mod$sample(
    data = data_list,
    seed = 123,
    chains = 4,
    refresh = 0,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000
)

print(fit)
```

Instruction: run the sampling command to draw posterior samples using NUTS. Use `print(fit)` for a quick console summary; use `fit$summary()` for a detailed table.

## Summarize & diagnose  {.smaller}

```{r}
library(posterior)
library(bayesplot)

draws <- as_draws_df(fit$draws())
summarize_draws(draws)
fit$cmdstan_diagnose()
```

Look for R-hat near 1, sufficient ESS, and no or few divergences.

Instruction: `summarize_draws()` gives posterior means/medians and intervals. `fit$cmdstan_diagnose()` reports sampler warnings — investigate any divergences or max tree depth hits.

## Workflow & recommended checks  {.smaller}

- Check sampling diagnostics: R-hat, ESS, divergences, tree depth
- Use PPCs to assess predictive fit
- Compare models with PSIS-LOO or cross-validation

## Visual diagnostics  {.smaller}

```r
# traceplots and density overlays
mcmc_trace(as.array(fit$draws(variables = c("beta","sigma"))))
mcmc_intervals(as.array(fit$draws(variables = "beta")), prob_outer=0.95)

# posterior predictive check
ppc_dens_overlay(y, as.matrix(fit$draws(variables = "y_rep")))
```

Instruction: use traceplots to check mixing and stationarity. Overlayed densities and PPCs help assess whether the model reproduces observed data patterns.

## Variational inference with ADVI {.smaller}

- **Variational inference (ADVI):** fast, approximate posterior (useful for large datasets or exploratory modeling).

```{r}
fit_vb <- mod$variational(data = data_list, seed = 123)
print(fit_vb)
```

## Optimization / MAP  {.smaller}

- **Optimization:** find posterior mode (MAP), useful for initialization, quick point estimate, or Laplace approximations.

```r
opt <- mod$optimize(data = data_list)
opt$summary()
```

## Pathfinder & Laplace approximation {.smaller}

- **Pathfinder:** fast method to find high-density proposals (check CmdStan/Pathfinder docs for availability).
- **Laplace approximation:** Gaussian approx around the mode using Hessian; often done after `optimize()`.

Instruction: these are approximate alternatives to full MCMC. Use them for rapid exploration but validate with sampling for final inference.

# brms: model building in R {.smaller}

## What is brms

```{r}
library(brms)
# simple linear model
fit_brm <- brm(y ~ x, data = df, chains = 2, cores=2, backend='cmdstanr', iter = 1000, refresh=0)
```

`brms` builds Bayesian models using formula syntax and can use Stan under the hood. Run this to fit a simple regression; use more chains and iterations for production use.

## brms: extract Stan code and data  {.smaller}

```r
standata <- standata(y ~ x, data = df)
stancode_text <- stancode(y ~ x, data = df)
```

You can save `stancode_text` to a `.stan` file and run with cmdstanr.

## brms: Linear mixed model {.smaller}

```{r}
set.seed(7)
G <- 10
group <- rep(1:G, each = N/G)
alpha_g <- rnorm(G, 1, 0.5)
y_lme <- alpha_g[group] + beta_true * x + rnorm(N, 0, sigma_true)
df_lme <- data.frame(x = x, y = y_lme, group = factor(group))
```

## brms: fit linear mixed model  {.smaller}

```{r}
fit_lme <- brm(y ~ x + (1|group), data = df_lme, cores = 2, chains=2, backend='cmdstanr', refresh=0)
fit_lme
```

Instruction: fit the mixed model with `brms`. `cores = 4` speeds up sampling by running chains in parallel; adjust to your machine.

## brms: quantile (LME quantile) example  {.smaller}

```{r}
# brms supports quantile-like models via the asymmetric Laplace family
fit_qr <- brm(bf(y ~ x + (1|group), quantile=0.5),
            data = df_lme,
            family = asym_laplace(),
            chains = 3,
            refresh = 0,
            prior = c(prior(normal(0,5), class = "b")),
            cores = 3)
print(fit_qr)
```

Note: check `brms` docs for the `asym_laplace` family and how to set the quantile parameter if needed.

## Summarize brms results  {.smaller}

```{r}
summary(fit_lme)
posterior_samples <- as_draws_df(fit_lme)
bayes_R2(fit_lme)
```

Discuss fixed effects, group SDs, R-hat, and ESS.

## bayesplot visualizations  {.smaller}

```r
library(bayesplot)
mcmc_trace(as.matrix(fit_lme))
mcmc_dens_overlay(as.matrix(fit_lme, pars = "b_x"))
pp_check(fit_lme, resp = "y")
```

Instruction: these visuals help you inspect mixing, posterior densities, and how well the model predicts observed data. Run them interactively to explore model fit.

## Other useful packages  {.smaller}

- **loo:** PSIS-LOO for model comparison (`loo` or `loo::loo()` / `brms::loo()`)
- **projpred:** projection predictive variable selection for Bayesian models

Instruction: use `loo` to compare models via approximate leave-one-out cross-validation; `projpred` helps with variable selection while preserving predictive performance.

# References  {.smaller}

- CmdStanR: https://mc-stan.org/cmdstanr/
- brms: https://paul-buerkner.github.io/brms/
- bayesplot: https://mc-stan.org/bayesplot/
- loo & projpred docs
